Univerzumok Típuselméletben

### 2021-04-08 efop_prez/prez.tex
translate efop prez to Hungarian

\title{Universes In Type Theory}
\date{April 8, 2021, EFOP Workshop}
\author{\normalsize{\vspace{-1em}\textbf{András Kovács}\footnote{The author was supported by the European Union, co-financed by the
European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}}
%% \institute{Eötvös Loránd University, Budapest}
\begin{document}
\maketitle

@@

\begin{frame}{Russell's paradox, or size issues}

There is no set $S$ in a consistent set theory such that $x \in S$ iff $x \notin x$.
\vspace{1em}

Alternate formulations:
\begin{itemize}
  \item Barber's paradox
  \item No set of all sets
\end{itemize}
\vspace{1em}

Related:
\begin{itemize}
  \item Real numbers are uncountable
  \item Gödel's first incompleteness theorem
  \item Undecidability of halting
\end{itemize}
\vspace{1em}

{\footnotesize (All instances of \href{https://arxiv.org/pdf/math/0305282.pdf}{Lawvere's fixed point theorem)}}
\end{frame}

@@

\begin{frame}{Solution with universes}

Solving Russell's paradox: set comprehension can only define a \textbf{subset} of a set.
\vspace{1em}

But sometimes we still want to quantify over \textbf{all sets} of
Zermelo-Fraenkel set theory.
\vspace{1em}

\textbf{Universe:} a set which includes every ZF set.
\vspace{1em}

Gödel: there is no universe inside ZF.
\vspace{1em}

We can assume more universes if we want to quantify over more sets.
\end{frame}

\begin{frame}[fragile]{Universes in type theory}

Type theories:
\begin{itemize}
  \item Alternative to set theory as mathematical foundation.
  \item Used in most general-purpose theorem provers.
  \item Can be used for proving and programming at the same time.
\end{itemize}

Universes come up in type theories as well.
\vspace{1em}

Every theorem prover has different universe-related features.
\vspace{1em}

We want to give a general semantics covering many different features.

@@

\begin{frame}[fragile]{Example}

@@

Identity function at all small sets:
\begin{verbatim}
  id : (A : Set₀) → A → A
  id A x = x
\end{verbatim}

We have $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$ for all $i$.
\vspace{1em}

Assuming $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$ implies contradiction (by Russell-like argument).
\vspace{1em}

@@

\begin{frame}[fragile]{Extensions}

The basic system can be tedious:
\begin{verbatim}
  id₀ : (A : Set₀) → A → A
  id₁ : (A : Set₁) → A → A
  ...
\end{verbatim}

In Coq/Agda/Lean, various extra features are used.
Example: universe polymorphism in Agda:
\begin{verbatim}
  id : (l : Level)(A : Set l) → A → A
  id l A x = x
\end{verbatim}

@@

\begin{frame}[fragile]{Design choices and variations}

\textbf{How many universes?} Agda/Coq: countably many.
\vspace{1em}

\textbf{Are universes totally ordered?} Agda/Coq: yes.
\vspace{1em}

\textbf{What kind of level polymorphism?}. Coq: bounded polymorphism. Agda: no
bounds allowed. Bounded example:
\begin{verbatim}
  myId : (i : ℕ) → i < 3 → (A : Set i) → A → A
  myId i p A x = x
\end{verbatim}

\textbf{What kind of operations are available on levels?} Agda is more liberal than Coq. Example:
\begin{verbatim}
  ℕtoLevel : ℕ → Level
\end{verbatim}

\textbf{Are universes cumulative} Agda: no. Coq: yes. Cumulativity: whenever \texttt{A : Set i}, we also have \texttt{A : Set (i + 1)}.

@@

\begin{frame}{Research goals}

We want to know that each point in the design space makes sense.
\vspace{1em}

Making sense:
\begin{itemize}
  \item Logical consistency.
  \item Is the type theory a proper programming language? Programs should compute to values
        and not get randomly stuck.
  \item Is proof checking decidable? (Not covered in current work).
\end{itemize}

Approach: use generic framework to cover as many features/variations as
possible. Prove that everything in the framework is sensible.

@@

\begin{frame}[fragile]{Features covered by the framework}

Universe levels may come from \textbf{any well-ordered set}, even transfinite:
\begin{verbatim}
  Set i : Set (i + 1) : ... : Set ω : Set (ω + 1) : ...
\end{verbatim}
\textbf{Quantification over levels}, \textbf{arbitrary computation on levels}.
\begin{verbatim}
   myId : (i : ℕ) → (j : ℕ) → (A : Set (i + j)) → A → A
   myId i j A x = x
\end{verbatim}
\textbf{Universes are cumulative}
\begin{verbatim}
   ℕ : Set₀
   ℕ : Set₁
   ...
\end{verbatim}

@@

\begin{frame}[fragile]{Implementation}

We define a family of type theories with general universe features.
\vspace{1em}

We prove all of them consistent by reducing all fancy features to
a single previously known feature, called \textbf{induction-recursion}.
\vspace{1em}

Induction-recursion is a type-theoretic analogue of assuming \textbf{Mahlo cardinals} in set theory.

@@

%% \begin{enumerate}
%%   \item Universe levels may come from any well-ordered set.
%%   \item Polymorphism with bounds is allowed.
%%   \item Universes are cumulative.
%%   \item Levels can be manipulated arbitrarily as program data (WIP).
%% \end{enumerate}

%% Things we need to assume in order to prove the framework sensible:
%% \begin{enumerate}
%%   \item \textbf{Two} universes.
%%   \item Types of certain infinitely branching trees (W-types).
%% \end{enumerate}

%% Potential further applications (in future work):
%% \begin{itemize}
%%   \item Information flow type systems (``secure'' and ``public'' levels).
%%   \item Staged compilation (``runtime'' and ``compile time'' levels).
%% \end{itemize}

%% \end{frame}

\begin{frame}{Publication}

@@

Submission under review for FSCD 2021.

@@

\begin{center} {\large Thank you!} \end{center}

>>>>

\title{Univerzumok Típuselméletben}
\date{2021 április 8., EFOP Workshop}
\author{\normalsize{\vspace{-1em}\textbf{Kovács András}\footnote{A szerzőt az Európai Unió támogatta, közös finanszírozásban az Európai Szociális Alappal (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}}
%% \institute{Eötvös Loránd University, Budapest}
\begin{document}
\maketitle

@@

\begin{frame}{Russell paradoxon és méret-problémák}

Nincs olyan $S$ halmaz egy konzisztens halmazelméletben, hogy $x \in S$ pontosan akkor, ha $x \notin
x$.
\vspace{1em}

Alternatív fogalmazás:
\begin{itemize}
  \item Borbély paradoxon
  \item Összes halmaz halmaza
\end{itemize}
\vspace{1em}

Kapcsolódó:
\begin{itemize}
  \item Valós számok nem megszámlálhatóak
  \item Gödel első nem-teljességi tétele
  \item Megállási probléma
\end{itemize}
\vspace{1em}

{\footnotesize (Általánosításuk: \href{https://arxiv.org/pdf/math/0305282.pdf}{Lawvere fixpont tétele)}}
\end{frame}

@@

\begin{frame}{Megoldás univerzumokkal}

Csak \textbf{létező halmaz részhalmazát} definiálhatjuk logikai formulával.
\vspace{1em}

De olykor szeretnénk kvantifikálni egy halmazelmélet \textbf{minden halmaza}
fölött.
\vspace{1em}

\textbf{Univerzum:} halmaz, ami tartalmazza egy halmazelmélet (pl. ZF) összes halmazát.
\vspace{1em}

Gödel, Russell: ZF-ben nincs univerzum, ami az összes ZF halmazt tartalmazza.
\vspace{1em}

Új univerzumokat axiómaként kell felvenni.
\end{frame}

\begin{frame}[fragile]{Típuselméletben}

Típuselméletek:
\begin{itemize}
  \item A halmazelméletek alternatívái.
  \item Tételbizonyító-rendszerek gyakori alapja.
  \item Bizonyításra és programozásra egyszerre alkalmas.
\end{itemize}

Formális vs. informális univerzum kezelés.
\vspace{1em}

Létező bizonyító-rendszerekben: univerzum megvalósítások sokasága kisebb-nagyobb eltérésekkel.
\vspace{1em}

Szeretnénk ezeket lefedni egy általános szemantikával.

@@

\begin{frame}[fragile]{Példa}

@@

Identitás függvény minden ``kis'' halmazon:
\begin{verbatim}
  id : (A : Set₀) → A → A
  id A x = x
\end{verbatim}

Minden $i$-re $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$.
\vspace{1em}

Ellentmondás lenne: $\mathsf{Set_i} : \mathsf{Set_{i}}$.
\vspace{1em}

@@

\begin{frame}[fragile]{Kiterjesztések}

Az alap rendszer vesződséges tud lenni:
\begin{verbatim}
  id₀ : (A : Set₀) → A → A
  id₁ : (A : Set₁) → A → A
  ...
\end{verbatim}

Coq/Agda/Lean-ben extra feature-ök vannak.
Pl. univerzum polimorfizmus Agda-ban:
\begin{verbatim}
  id : (l : Level)(A : Set l) → A → A
  id l A x = x
\end{verbatim}

@@

\begin{frame}[fragile]{Variációk}

\textbf{Mennyi univerzum?} Agda/Coq: megszámlálható.
\vspace{1em}

\textbf{Univerzumok teljesen rendzettek?} Agda/Coq: igen.
\vspace{1em}

\textbf{Milyen univerzum polimorfizmus?}. Coq: korlátos polimorfizmus. Agda: nincsenek
korlátok. Példa korláttal:
\begin{verbatim}
  myId : (i : ℕ) → i < 3 → (A : Set i) → A → A
  myId i p A x = x
\end{verbatim}

\textbf{Milyen műveletek vannak univerzum-szinteken?} Az Agda liberálisabb a Coq-nál:
\begin{verbatim}
  ℕtoLevel : ℕ → Level
\end{verbatim}

\textbf{Kumulativitás.} Agda: nincs. Coq: van. Kumulativitás: ha \texttt{A : Set i}, akkor \texttt{A
  : Set (i + 1)}.

@@

\begin{frame}{Kutatási célok}

Szeretnénk belátni, hogy a különféle variációk mind jól viselkednek.
\vspace{1em}

Jól viselkedés:
\begin{itemize}
  \item Logikai konzisztencia.
  \item Igaz, hogy a típuselmélet egy programozási nyelv? Azaz: programok le kell, hogy
        fussanak elakadás nélkül. (Nincs jelenlegi munkában)
  \item Eldönthető a bizonyítások validálása? (Nincs jelenlegi munkában)
\end{itemize}

Egy általános rendszert használunk, ahol sok feature modellezhető, és
egyszerre bizonyítjuk az összes konzisztenciáját.

@@

\begin{frame}[fragile]{Amit a rendszerünk lefed}

Univerzum-szintek \textbf{bármilyen jól-rendezett halmazt} alkothatnak.
\begin{verbatim}
  Set i : Set (i + 1) : ... : Set ω : Set (ω + 1) : ...
\end{verbatim}
\textbf{Szintek fölötti kvantifikáció}, \textbf{tetszőleges számítások szinteken}.
\begin{verbatim}
   myId : (i : ℕ) → (j : ℕ) → (A : Set (i + j)) → A → A
   myId i j A x = x
\end{verbatim}
\textbf{Kumulativitás}
\begin{verbatim}
   ℕ : Set₀
   ℕ : Set₁
   ...
\end{verbatim}

@@

\begin{frame}[fragile]{Megvalósítás}

Visszavezetjük az összes feature-t egy már ismert feature-re,
az \textbf{induktív-rekurzív} definíciókra.
\vspace{1em}

Az indukció-rekurzió egy típuselméleti változata a halmazelméleti \textbf{Mahlo kardinálisoknak}.

@@

\begin{frame}{Publikáció}

@@

Elbírálás alatt FSCD 2021 konferenciára.

@@

\begin{center} {\large Köszönöm a figyelmet!} \end{center}
